---
title: "Conclusion"
format:
  html: 
    toc: false
    link-external-icon: true
    link-external-newwindow: true
    link-external-filter: ^(?:http:|https:)\/\/www\.openstatprog\.com
---

```{r include = FALSE}
source(here::here("R", "buttons.R"))
```

By completing this module, you've built a solid foundation in understanding two more of R's essential data
structures:

- Lists
- Data Frames

## Hand-In Assignment

::: practice

## Lists & Data Frames Capstone Assignment

Complete each step below in your own R session, then submit your work (see "What to Hand In" below).

1. **Create a list.** Build a list called `book` describing a book of your choice, with named slots for `title`
   (character), `pages` (numeric), and `finished` (logical, whether you've read it).
1. **Access and explain.** Access the `pages` slot of `book` two different ways (e.g., using `$` and `[[]]`). In a
   comment, explain why `book["pages"]` would *not* give you the same kind of result as `book[["pages"]]` or
   `book$pages`.
1. **Create a data frame.** Build a data frame called `books` with (at least) 4 rows and 3 columns: `title`
   (character), `pages` (numeric), and `finished` (logical). You can make up the data.
1. **Select and modify.** Using indexing (your choice of list-style or matrix-style), select all books in `books`
   where `pages` is greater than some threshold of your choosing, and change their `finished` value to `TRUE`.
1. **Conceptual explanation.** In a comment, explain in your own words how `books` (a data frame) is related to
   `book` (a list). Specifically, what makes a data frame a *special case* of a list rather than something entirely
   different.

### What to Hand In

Save all of your commands, in the order you ran them, in a single plain-text or R script file (e.g.
`yourname_lists-data-frames.R`), including both required comments. Submit the file however your instructor has
requested.

A teacher reviewing your submission should be able to see: a correctly constructed list, two working (and correctly
explained) list-access methods, a correctly constructed data frame, correct conditional selection/modification, and a
clear explanation of the list–data-frame relationship in your own words.

:::

## Learning Goals

Now that you've completed this module, you should be able to:

### Lists

1. Describe the characteristics of an R list.
1. Describe important use-cases for lists, and explain why lists are the best choice in these situations.
1. Create new lists in R.
1. Access and modify list elements using different methods and explain the difference between the available
   selection operators.

### Data Frames

1. Describe the characteristics of data frames.
1. Explain important use-cases for data frames, and explain why data frames are the best choice in these situations.
1. Explain how data frames and lists are related.
1. Explain the differences between data frames and matrices.
1. Create new data frames in R.
1. Use list-style and matrix-style indexing to access and modify the elements of a data frame.
1. Explain the different effects of the available selection operators.
1. Explain how recycling works with data frames.

## Additional Reading

For more information on the topics we've covered in this module, the following resources are excellent starting
points.

- [Lists](https://r4ds.hadley.nz/vectors.html#lists) and [Data Frames](https://r4ds.hadley.nz/data-structures.html)
  chapters from Hadley Wickham's *R for Data Science*.
- [Tibbles](https://r4ds.hadley.nz/data-import.html) for more on the tibble variant briefly previewed in this module.

### Advanced Reading

For a much more technical deep-dive into the topics we've covered in this module, check out the following chapter
from Hadley Wickham's [Advanced R](https://adv-r.hadley.nz/index.html).

- [Subsetting](https://adv-r.hadley.nz/subsetting.html)

## Next Steps 

Now that you're comfortable with lists and data frames, you're ready to move on to the last data structure in this
series: factors, which R uses to represent categorical data.

```{r}
#| echo: false
#| results: asis

next_module_button("https://www.openstatprog.com/factors/index.html")
```
