---
title: "Creating Data Frames"
---

::: {.callout-note}
This page is hands-on: constructing data frames with columns of different types.
:::

We create data frames using the `data.frame()` function. This function takes vectors as arguments, and each vector
becomes a column in the resulting data frame. 

In the following code, we create a data frame called `d1` that contains 10 rows and 3 columns.

1. The first column is the integer sequence from 1 to 10
1. The second column alternates between -1 and 1
1. The third column contains the sequence $\{0.1, 0.2, \ldots, 1.0\}$.

```{r}
(d1 <- data.frame(1:10, c(-1, 1), seq(0.1, 1, 0.1)))
```

Since we did not name the arguments to `data.frame()` when creating `d1`, R generated default (ugly) column names by
parsing the argument values. 

```{r}
colnames(d1)
```

We can assign our own column names by naming the arguments to `data.frame()`.

```{r}
# Generate a version of d1 with better column names
(d2 <- data.frame(x = 1:10, y = c(-1, 1), z = seq(0.1, 1, 0.1)))
```

We can also assign column names to an existing data frame using the `names()` or `colnames()` functions.

```{r}
names(d1) <- c("foo", "bar", "baz")
d1

colnames(d1) <- c("alice", "bob", "suzy")
d1
```

One of the greatest strengths of data frames is their ability to store heterogeneously typed columns. For example, in
the following code, we create a data frame called `d3` that comprises three uniquely typed columns.

a. A logical vector containing 10 randomly sampled boolean values.
a. A character vector containing 10 random selections from the set {"foo", "bar"}.
a. A numeric vector containing 10 values randomly sampled from the interval [0, 1].

```{r}
(d3 <- data.frame(a = sample(c(TRUE, FALSE), 10, replace = TRUE),
                  b = sample(c("foo", "bar"), 10, replace = TRUE),
                  c = runif(10)))
```

If we want to quickly generate an empty data frame, we can supply a matrix as the sole argument to `data.frame()`.

```{r}
(d4 <- data.frame(matrix(NA, 10, 3)))
```

::: practice

## Practice

Create a data frame containing 7 observations of 3 variables.

- Use a different type for each of the variables.

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.solution >}}

There are an infinite number of correct solutions to this problem (literally). The following is one possibility.

```{r}
df <- data.frame(text = rep("text", 7),
                 integer = sample(1:10, 7, replace = TRUE),
                 complex = complex(7, 1:5, 1:3)
                 )
df
```

::::

:::

## Quick Check

- What determines the column names when you don't explicitly name the arguments to `data.frame()`?
- Can two columns in the same data frame have different types? Can they have different lengths? Why or why not?

```{webr-r}
```
