---
title: "Data Frames"
---

::: {.callout-note}

## Learning Outcomes

By the end of this tutorial, you will be able to:

- **Describe** the characteristics of data frames.
- **Explain** important use-cases for data frames, and explain why data frames are the best choice in these
  situations.
- **Explain** how data frames and lists are related. 
- **Explain** the differences between data frames and matrices.
- **Create** new data frames in R.
- **Use** list-style and matrix-style indexing to access and modify the elements of a data frame, and explain the
  different effects of the available selection operators. 
- **Explain** how recycling works with data frames.

:::

Data frames are one of the most useful data structures in R, at least when it comes to real-world data analysis. A data
frame is a two-dimensional, tabular object wherein the columns must be the same length, but they can contain different
types of data. This structure closely mirrors the layout of rectangular, spreadsheet-style datasets, so data frames are
the natural choice for storing and manipulating real-world datasets in most data analytic projects — anywhere you'd
reach for a spreadsheet, a data frame is usually the R equivalent.

## Data Frames Are Actually Lists

Interestingly, data frames are actually implemented as a special case of lists. Each column in a data frame is an
element of a list, but all of these list elements must be vector-like objects (not necessarily atomic vectors, though)
with equal length. In other words, a data frame is just a list where each slot contains a vector representing one column
in the data frame, but with additional constraints and attributes that allow it to behave like a 2 dimensional table.

```{r}
d1 <- data.frame(a = sample(c(TRUE, FALSE), 10, replace = TRUE),
                 b = sample(c("foo", "bar"), 10, replace = TRUE),
                 c = runif(10)
                 )
d2 <- data.frame(a = 1:10, b = c(-1, 1), c = seq(0.1, 1, 0.1))
d3 <- data.frame(x = -5:4, y = c(1, 10), z = seq(1, 1.5, length.out = 10))

# d1, d2, and d3 are all data frame.
is.data.frame(d1)
is.data.frame(d2)
is.data.frame(d3)

# But they're also lists
is.list(d1)
is.list(d2)
is.list(d3)
```

All data frames are lists, but not all lists need be data frames.

```{r}
l1 <- list()
l2 <- list(a = sample(c(TRUE, FALSE), 10, replace = TRUE),
           b = sample(c("foo", "bar"), 10, replace = TRUE),
           c = runif(10)
           )
l3 <- list(d2, d3)

# l1, l2, and l3 are all lists
is.list(l1)
is.list(l2)
is.list(l3)

# But they're not data frames
is.data.frame(l1)
is.data.frame(l2)
is.data.frame(l3)
```

## Data Frames Are Not Matrices

As you have probably surmised, data frames are not matrices. Although data frames and matrices share many superficial
similarities, they're almost entirely unrelated types of objects.

```{r}
is.matrix(d1)
is.matrix(d2)
is.matrix(d3)
```

Consequently, you cannot apply matrix operations (such as matrix multiplication) to a data frame.

```{r error = TRUE}
d2 %*% t(d3)
```

If you wanted to analyze a data frame like a matrix, you could *type cast* the data frame into a matrix using the
`as.matrix()` function.

```{r}
as.matrix(d2) %*% t(as.matrix(d3))
```

You should be carefully about any type conversions. If you're data frame contains mixed column types, R will coerce every
column into some compatible type, the same coercion rule you learned about back in the Vectors tutorial, just
applied across an entire data frame at once.

```{r}
(m1 <- as.matrix(d1))
typeof(m1)

(d4 <- data.frame(x = rep(c(TRUE, FALSE), 3), y = runif(6)))
(m4 <- as.matrix(d4))
typeof(m4)
```

## Quick Check

- In one sentence: how is a data frame related to a list?
- Why can't you do `d2 %*% t(d3)` directly on two data frames?
- If a data frame has one text column and one numeric column, what happens if you convert it to a matrix with
  `as.matrix()`? Why?
