---
title: "Manipulating Data Frames"
knitr:
  opts_chunk:
    warning: true
---

## Accessing Data Frame Elements

Because data frames are just lists, we can access a data frame's columns using the same methods we would use for lists.
To access a single column by name, the most efficient method is typically the `$` operator.

```{r}
d1 <- data.frame(
  a = sample(c(TRUE, FALSE), 10, replace = TRUE),
  b = sample(c("foo", "bar"), 10, replace = TRUE),
  c = runif(10)
)

d1
d1$b
```

We can also use the single, `[]`, or double, `[[]]`, square bracket operators. As with lists, these operators differ in
two respects: how many columns they can select and how they format the resulting selection.

- `[]` Can select multiple columns and always returns a data frame.
- `[[]]` Can select only one column and returns the column contents as a vector (or whatever type of object the column
contained).

```{r}
# Return a one-column data frame comprising the 'b' column from d1
d1["b"]

# The same as above, but using the column index instead of the column name
d1[2]

# Return a two-column data frame comprising the 'a' and 'b' columns from d1
d1[c("a", "b")]

# The same as above, but using the column indices instead of the column names
d1[1:2]

# Return the 'b' column from d1 as a character vector
d1[["b"]]

# The same as above, but using the column index instead of the column name
d1[[2]]
```

### Matrix-Style Selection

Data frames also support matrix-style subsetting, where we define the selection by specifying both the row and column
indices.

```{r}
# Return the first two rows of columns `b` and `c`
d1[1:2, 2:3]

# Return all rows from columns `a` and `b`
d1[, 1:2]

# Return the second and third rows of all columns
d1[2:3, ]
```

### Quirks of Matrix-Style Subsetting

In most cases, matrix style subsetting behaves the same was as the `[]` list-style operator. You can select any number
of columns, and the selection is returned as a data frame, but there is one important exception. If you select a single column
using matrix-style subsetting, the selection will be converted to a vector.

```{r}
d1[, 1]
```

::: going-further

If you want to create a one-column data frame by extracting a single column from another data frame, you should probably
use the `[]` list-style subsetting operator. That being said, if you really want to use matrix-style subsetting, you can
achieve the desired effect by adding the `drop = FALSE` argument inside the brackets.

```{r}
d1[, 1, drop = FALSE]
```

I wouldn't recommend using this pattern since it produces non-idiomatic R code (i.e., by convention, we should only
specify function arguments inside the parentheses of a function call). You might come across this syntax in the wild,
however, so it's worth understanding what's happening.

:::

### Convenience Functions

Calling the `head()` function on a data frame returns the first `n` rows of the data frame.

```{r}
head(d1, 3)
```

Likewise, the `tail()` function returns the final `n` rows of the data frame.

```{r}
tail(d1, 3)
```

As with vectors, providing a negative value for the `n` argument returns the complementary number of rows from either
the top or bottom of the data frame.

```{r}
head(d1, -3)
tail(d1, -3)
```

::: practice

The `iris` dataset is provided by Base R and available in any R session.

```{r}
head(iris, 10)
```

Use the `iris` dataset to complete the following tasks.

:::: {.panel-tabset}

## {{< var tab_title.question >}}

1. Use three different methods to extract the `Petal.Length` column.
   - Return the results as vectors.
1. Extract the `Petal.Width` column.
   - Return the result as a one-column data frame.
1. Extract the first 20 rows of the `Sepal.Length` and `Sepal.Width` columns.
   - Return the result as a $20 \times 2$ data frame.
1. Extract the final 10 rows of the data frame.
   - Return the result as a $10 \times 5$ data frame.

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

**Q1**

```{r}
iris$Petal.Length
iris[["Petal.Length"]]
iris[, "Petal.Length"]
```

**Q2**

```{r}
iris["Petal.Width"]
```

**Q3**

```{r}
iris[1:20, c("Sepal.Length", "Sepal.Width")]
```

**Q4**

```{r}
tail(iris, 10)
```

::::

:::

## Modifying Data Frame Elements

At this point, I hope it comes as no surprise to hear that we can overwrite the columns of a data frame using the same
procedures that we use to modify list slots. When modifying one column at a time, we directly apply the intuitive
operations.

```{r}
# View the original data frame
d1

# Modify some columns
d1$a <- LETTERS[1:10]
d1[[2]] <- rnorm(10)
d1["c"] <- rep(c(TRUE, FALSE), each = 5)

# View the modified data frame
d1
```

When modifying multiple columns with the `[]`, operator, it's best to supply the replacement values as a data frame or
list with the same size as the selected columns. For example, below we replace the first two columns of `d1` with an
equivalently sized data frame.


```{r}
d1[1:2] <- data.frame(1:10, 1:2)
d1
```

Here, we replace the `a` and `c` columns in `d1` with an equivalently sized list.

```{r}
d1[c("a", "c")] <- list(rnorm(10), runif(10))
d1
```

### Matrix-Style Selection

If we only want to replace part of a column, we can use matrix-style selection to choose the target cells. Below, we
first overwrite the first five rows in column `b`. We then overwrite rows 3 through 6 in columns `a` and `c`.

```{r}
d1[1:5, 2] <- 41:45
d1

d1[3:6, c("a", "c")] <- list(-99, 888)
d1
```

In the second command, notice the pattern in which the contents of the list are recycled to fill the target cells. The
recycling rules for data frames can be a little tricky.

## Recycling

When the replacement size doesn't match the selection size, R will use recycling to resolve the discrepancy, as with
vectors and matrices. However, it's not always easy to predict how the replacement will behave.

If we supply a single vector as the replacement data, the recycling operates the same way it does for matrices. For
example, in the following code, we replace the first two columns of `d1` by recycling the vector `1:5`.

```{r}
d1[1:2] <- 1:5
d1
```

Nothing new here. We simply re-use the elements of the replacement vector, `1:5`, until each selected cell is
overwritten. Things get more complicated with we supply the replacement data as a list or data frame, though. Below, we
want to replace the `a` and `c` columns in `d1` with the contents of a list, but the vectors therein aren't long enough,
so they will need to be recycled.

```{r}
d1[c("a", "c")] <- list(c("yes", "no"), 8:12)
d1
```

Note that each list slot acts as the replacement for one of the selected columns. So, the recycling happens within
columns: the contents of each list slot are recycled independently to resolve the respective size discrepancy. In this
case,

- The vector `c("yes", "no")` is recycled five times to match the length of column `a`.
- The vector `8:12` only needs to be recycled twice.

### Restrictions

As with matrices, R is oddly specific (in a slightly different way) about the kinds of size discrepancies it will
automatically resolved when modifying data frames. When all lengths align cleanly (as in the examples above), R will
apply recycling silently (i.e., without any messages or warnings).

**Ideal**

- All replacement lengths $\leq$ Respective selection lengths
- All replacement lengths evenly divide their respective selection lengths
- Number of replacement list slots = Number of columns selected
- Number of replacement data frame columns = Number of columns selected

For several other types of size discrepancy, R will apply recycling with a warning.

**Acceptable**

- Replacement length > Selection length
- Replacement length cleanly divides the selection length
- Replacement length exceeds selection length
- Replacement list contains more slots than columns selected
- Replacement data frame contains more slots than columns selected

In one specific situation, R will throw an error and refuse to implement the replacement.

**Invalid**

- Replacement length does not evenly divide the selection length.

#### Examples of Acceptable Replacement

When working with data frames, R is quite permissive with abuses of recycling. So, all of the following examples
represent legal uses of replacement that R will run without errors, but these examples will all trigger warnings.

Replace the first two columns of `d1` by using the first 20 elements from the vector `100:500`.

```{r}
#| warning: true

d1[1:2] <- 100:500
d1
```

Replace the `a` and `c` columns in `d1` with a list containing vectors that are too long.

```{r}
#| warning: true

d1[c("a", "c")] <- list(rnorm(100), runif(100))
d1
```

Replace the `a` and `c` columns in `d1` with the first two slots in a length-3 list.

```{r}
#| warning: true

d1[c("a", "c")] <- list(1, 2, 3)
d1
```

Replace the `a` and `c` columns in `d1` with the first two columns of a three-column data frame.

```{r}
#| warning: true

d1[c("a", "c")] <- data.frame("foo", "bar", "baz")
d1
```

#### Examples of Invalid Replacement

For the most part, R will only complain when the replacement length does not cleanly divide selection length. So, the
following two examples represent some of the few situations in which R will completely refuse to apply our requested
replacement.

Here, we try to replace the first two columns of `d1` (i.e., 20 total cells) with the length-three vector `1:3`.

```{r}
#| error: true

d1[1:2] <- 1:3
```

This time, we make essentially the same mistake, but now the incompatible vectors live in a list. We're trying to
replace the `a` and `c` columns in `d1` with a list containing a length-four vector and a length-eight vector.

```{r}
#| error: true

d1[c("a", "c")] <- list(letters[4], 1:8)
```

## Adding Columns

As with lists, we can add new columns to an existing data frame using the `$` or `[[]]` operators.

```{r}
# Create a baseline data frame
(d2 <- data.frame(a = rep(0, 10)))

# Various ways of adding new single columns
d2$b <- letters[1:10]
d2[["c"]] <- runif(10)
d2[[4]] <- rnorm(10)
d2
```

As you may expect, R will use recycling when we specify too few elements for the new column.

```{r}
d2["alice"] <- c(TRUE, FALSE)
d2
```

We can add multiple columns using the `[]` operator.

```{r}
d2[6:7] <- rnorm(20)
d2[c("foo", "bar")] <- list(TRUE, FALSE)
d2
```

::: practice

Run the following code to create an empty data frame containing 10 observations of the 3 variables: `a`,`b`, `c`. Then
populate the data frame as described below.

1. Fill column `a` with the integer sequence from -9 to 0.
   - Use the column name to assign the new values.
1. Fill column `b` with the even integers between 1 and 20 (inclusive).
   - Use the numeric column index to assign the new values.
1. Replace the odd rows in column `c` with the odd integers between 11 and 20 (inclusive).
   - Do not overwrite the missing values in the even rows. 

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
df <- data.frame(
  a = rep(NA, 10),
  b = rep(NA, 10),
  c = rep(NA, 10)
)
```

## {{< var tab_title.solution >}}

```{r}
df <- data.frame(a = rep(NA, 10), b = rep(NA, 10), c = rep(NA, 10))
df

df$a <- -9:0
df[2] <- seq(2, 20, 2)
df[seq(1, 9, 2), "c"] <- seq(11, 19, 2)
df
```

::::

:::

::: knowledge-check

- If `d1` is a data frame, why does `d1[2]` return a one-column data frame, while `d1[[2]]` returns a vector?
- Suppose you try to replace two columns in a data frame with the contents of a three-slot list.
  - Which list slots (if any) does R use to fill the columns?
  - Will R raise an error?
  - Will R show a warning?

:::
