---
title: "Relation to Lists"
---

::: {.callout-important}

Data frames are somewhat strange objects that can cause a lot of confusion. To new users, data frames usually feel more
similar to matrices than lists, but the opposite is actually true. The underlying implementation of data frames
doesn't match their apparent tabular structure. This page compares data frames to lists and matrices. Understanding
these similarities and differences will help you use data frames more effectively and avoid many common mistakes.

:::

Did you notice the similarity between the code that we used to create data frames on the last page and the code that we
used to create lists in the last tutorial?

```{r}
# Create a data frame
someDataFrame <- data.frame(
  a = rnorm(10),
  b = 1:10,
  c = letters[1:10]
)

# Create a list
someDataFrame <- list(
  a = rnorm(10),
  b = 1:10,
  c = letters[1:10]
)
```

If so, you were right to see some parallels: lists and data frames are actually the same thing!

## Data Frames Are Actually Lists

Data frames are actually implemented as a special kind of list. Each column in a data frame is an element of a list, but
all of these list elements must be vector-like objects (not necessarily atomic vectors, though) with equal length. In
other words, a data frame is just a list where each slot contains a vector representing one column in the data frame,
but with additional constraints and attributes that allow the data frame to behave like a 2 dimensional table.

```{r}
d1 <- data.frame(
  a = sample(c(TRUE, FALSE), 10, replace = TRUE),
  b = sample(c("foo", "bar"), 10, replace = TRUE),
  c = runif(10)
)
d2 <- data.frame(a = 1:10, b = c(-1, 1), c = seq(0.1, 1, 0.1))
d3 <- data.frame(x = -5:4, y = c(1, 10), z = seq(1, 1.5, length.out = 10))
```

For example, `d1`, `d2`, and `d3` are all data frames.

```{r}
is.data.frame(d1)
is.data.frame(d2)
is.data.frame(d3)
```

But all three are also lists.

```{r}
is.list(d1)
is.list(d2)
is.list(d3)
```

### Not all Lists are Data Frames

All data frames are lists, but the inverse is not always true. Some lists are not data frames.

```{r}
l1 <- list()
l2 <- list(
  a = sample(c(TRUE, FALSE), 10, replace = TRUE),
  b = sample(c("foo", "bar"), 10, replace = TRUE),
  c = runif(10)
)
l3 <- list(d2, d3)
```

As you can see below, `l1`, `l2`, and `l3` are all lists, but they're not data frames.

```{r}
is.list(l1)
is.list(l2)
is.list(l3)

is.data.frame(l1)
is.data.frame(l2)
is.data.frame(l3)
```

### List $\leftrightarrow$ Data Frame Conversion

We can use the `as.list()` type conversion function to convert a data frame to a list. For example, in the following
code, we convert `d1` into a list, `l1a`.

```{r}
(l1a <- as.list(d1))
is.list(l1a)
is.data.frame(l1a)
```

Likewise, we can use the `as.data.frame()` to convert a list to a data frame. Below, we invert the conversion from
above.

```{r}
(d1a <- as.data.frame(l1a))
is.data.frame(d1a)
```

In most situations, these type conversions do not destroy an information in the underlying data. Below, we see that the
converted-and-back-converted version of the data frame, `d1a`, is identical to the original data frame, `d1`.

```{r}
identical(d1, d1a)
```

## Data Frames Are Not Matrices

As you have probably surmised, data frames are not matrices. Although data frames and matrices share many superficial
similarities, they're almost entirely unrelated types of objects.

```{r}
is.matrix(d1)
is.matrix(d2)
is.matrix(d3)
```

Consequently, you cannot apply matrix operations (such as matrix multiplication) to a data frame.

```{r error = TRUE}
d2 %*% t(d3)
```

If you wanted to analyze a data frame like a matrix, you could *type cast* the data frame into a matrix using the
`as.matrix()` function.

```{r}
as.matrix(d2) %*% t(as.matrix(d3))
```

You should be carefully about any type conversions, though. If your data frame contains mixed column types, R will
coerce every column into some compatible type by applying the same coercion rule you learned about in the [**Vectors &
Matrices**](https://openstatprog.com/vectors-matrices/) module.

```{r}
(m1 <- as.matrix(d1))
typeof(m1)

(d4 <- data.frame(x = rep(c(TRUE, FALSE), 3), y = runif(6)))
(m4 <- as.matrix(d4))
typeof(m4)
```

:::knowledge-check

- In one sentence, explain how a data frame is related to a list.
- if `a` and `b` are both data frames, why will `a %*% t(b)` produce an error?
- If a data frame has one text column and one numeric column, what happens if you convert it to a matrix with
  `as.matrix()`? Why?

:::
