---
title: "Tibbles: A Modern Data Frame"
knitr:
  opts_chunk:
    eval: true
---

::: {.callout-note}
The goal is to recognize tibbles and describe how they relate to ordinary data frames, not to master a new syntax. None of the code on this page needs to run for you to get the point. Read it, and try it yourself later once you've installed the **tibble** package (or the tidyverse, which includes it).
:::

::: {.callout-tip}

## Optional Page

This page is a brief preview, not a required part of the module's learning outcomes. Feel free to skim it now and
come back later once you've had more practice with ordinary data frames.

:::

As you continue working with R, especially once you start using packages from the [tidyverse](https://www.tidyverse.org/)
(like **dplyr** or **ggplot2**), you'll frequently encounter a variant of the data frame called a **tibble**. A tibble
is provided by the **tibble** package, and it's designed to behave like a data frame in almost every way, while
smoothing over a few behaviors that regularly surprise beginners.

## Creating a Tibble

We create tibbles with the `tibble()` function, which works almost identically to `data.frame()`.

```{r}
library(tibble)

t1 <- tibble(x = 1:5, y = c("a", "b", "c", "d", "e"))
t1
```

If you ran this yourself, you'd notice the printed output already looks a little different from an ordinary data
frame. Tibbles print the number of rows and columns, and show each column's type right under its name.

## How Tibbles Differ From Data Frames

Tibbles are still data frames underneath (in fact, `is.data.frame()` returns `TRUE` for a tibble). They change a
few specific behaviors, not the fundamental structure:

```{r}
is.data.frame(t1)
class(t1)
```

- **Tibbles never simplify on selection.** Remember how selecting a single column from a data frame using
  matrix-style indexing (`d1[, 1]`) automatically "dropped" the result down to a plain vector? Tibbles don't do that.
  `t1[, 1]` always returns a one-column tibble, never a bare vector. This removes a common source of unexpected
  behavior in longer scripts.
- **Tibbles don't do partial column-name matching.** With an ordinary data frame, `df$he` might accidentally match a
  column called `height` if there's no exact match. Tibbles require an exact match, which helps catch typos early.
- **Tibbles print more safely.** Printing a large ordinary data frame dumps every row to your console. Tibbles print
  only the first several rows by default, along with each column's type — much friendlier for big datasets.

```{r}
# Compare: selecting one column with matrix-style indexing
d1 <- data.frame(x = 1:5, y = letters[1:5])

class(d1[, 1])  # a data frame "drops" to a vector
class(t1[, 1])  # a tibble does not drop
```

## Converting Between the Two

You can convert freely between ordinary data frames and tibbles.

```{r}
# Convert a data frame to a tibble
as_tibble(d1)

# Convert a tibble to an ordinary data frame
as.data.frame(t1)
```

## Quick Check

- Is a tibble a completely different data structure from a data frame, or a variant of one?
- Name one specific behavior where tibbles differ from ordinary data frames, and explain why that difference exists.
