---
title: "Tibbles: A Modern Data Frame"
knitr:
  opts_chunk:
    eval: true
    warning: true
---

::: going-further

## Optional Content

This page is a brief preview of an advanced topic that you'll encounter in later modules, not a required part of this
module's learning outcomes. Feel free to skim the following information now and come back later once you've had more
practice with ordinary data frames.

:::

As you continue working with R---especially once you start using packages from the [tidyverse](https://www.tidyverse.org/)
like **dplyr** and **ggplot2**---you'll frequently encounter a special kind of data frame called a **tibble**. This page
provides a brief introduction to these very handle data objects.

## Creating Tibbles

Tibbles are provided by the **tibbles** package, not Base R. So, before you can use tibbles, you must first install the
**tibble** package.

```{r}
#| eval: false

install.packages("tibble")
```

We create tibbles with the `tibble()` function, which works (mostly) like `data.frame()`.

```{r}
library(tibble)

t1 <- tibble(
  x = rnorm(10),
  y = sample(c("foo", "bar", "baz"), 10, TRUE),
  z = x^2
)
t1
```

Notice how the printed output looks a little different from an ordinary data frame. Tibbles print the number of rows and
columns and show each column's type under the column name. When printing large tibbles, the rows and/or columns will be
truncated into a sensibly sized summary.

```{r}
as_tibble(
  matrix(
    rnorm(10000),
    ncol = 50,
    dimnames = list(NULL, paste0("x", 1:50))
  )
)
```

This bespoke print method is one difference between tibbles and traditional data frames, but there are several others
(including a few very handy improvements).

## How Tibbles Differ from Data Frames

In many ways, tibbles *don't* differ from data frames. For the most part, tibbles are meant to be a drop-in replacement
for traditional data frames, so they don't change much about the core data frame features (after all, data frames are
already very useful). As far as R is concerned, tibbles are still data frames underneath.

```{r}
is.data.frame(t1)
```

Since they share the same underlying structure, you can seamlessly convert data frames to tibbles and visa-versa.

```{r}
# Convert a tibble to an ordinary data frame
(d1 <- as.data.frame(t1))

# Convert a data frame to a tibble
as_tibble(d1)
```

The goal of tibbles is to smooth over some of the confusing, frustrating, or surprising behavior of traditional data
frames. Below, we highlight a few of the most useful differences.

### Tibbles Don't Do Surprise Simplification

Remember how selecting a single column from a data frame using matrix-style indexing (`d1[, 1]`) automatically
simplifies the result to a vector? I certainly hope so; we just covered that on the last page. Anyway, tibbles don't do
that.

Subsetting a tibble using matrix-style indexing always returns a tibble, regardless of how many rows or columns are
extracted.

```{r}
t1[1:2, 1:2]
t1[, 1]
t1[1, ]
```

### Tibbles Don't Do Partial Name Matching

We haven't explicitly addressed it yet, but many R commands use "partial matching" to resolve your inputs. For example,
we don't actually have to write out the full word `digits` to specify the corresponding argument in `round()`. We only
need to provide enough of the initial letters to disambiguate the intended argument name.

```{r}
round(digits = 2, 3.14159)
round(dig = 2, 3.14159)
round(d = 2, 3.14159)
```

Ordinary data frames use the system to match column names. So, to match the column `foo` in the following data frame, we
only need to specify one letter.

```{r}
d1 <- data.frame(
  foo = rnorm(10),
  bar = runif(10),
  baz = rbinom(10, 1, 0.5)
)

d1$f
```

To match the `bar` or `baz` columns, however, we need to give the full column name, since these two names only differ in
their final letter.

```{r}
#| error: true

d1$ba
```

Usually, this behavior doesn't cause problems, but it creates unnecessary ambiguity, and increases the opportunities for
bugs. Consider the following hypothetical data frame.

```{r}
myData <- data.frame(
  female = sample(0:1, 20, TRUE),
  malevolent = sample(0:1, 20, TRUE),
  score = rnorm(20)
)

myData
```

Suppose the `female` column contains a dummy code representing biological sex. Further suppose that I want to select the
`female` column, but I misremember its name as `male`, so I run the following command.

```{r}
myData$male
```

Oops...hope I'm paying extra close attention. Otherwise, as I continue with my analysis, the cases that I think are
biological males are actually the cases that have been flagged as "malevolent" by the dummy code that I accidentally
selected via silent partial matching.

Tibbles don't allow partial matching for columns names, so this type of mistake cannot happen when working with a tibble.

```{r}
#| error: true

myData2 <- as_tibble(myData)

myData2$male
```

### Tibbles Dramatically Restrict Recycling

Tibbles do away with essentially all of the convoluted recycling rules we've discussed for other type of R objects.
Basically, we're only allowed to recycle length-one vectors. So, the following commands are all accepted.

```{r}
myData2$female <- 42
myData2

myData2[1:6, "score"] <- pi
myData2

myData2[1:2] <- list("foo", FALSE)
myData2
```

These commands, on the other hand, will all fail because we're trying to recycle vectors with more than one element.

```{r}
#| error: true

myData2$female <- 40:42
myData2[1:6, "score"] <- c(pi, NA)
myData2[1:2] <- list("foo", c(FALSE, TRUE))
```

::: knowledge-check

- How are tibbles and data frames related?
- What's the purpose of tibbles, when we already have perfectly serviceable data frames?
- Name one specific behavior where tibbles differ from ordinary data frames, and explain the pros and cons of that
difference.

:::

::: {.callout-tip}

## Ready to Continue?

You're ready to progress if you can:

- Create a data frame with columns of different types.
- Explain how a data frames are related to lists and matrices.
- Use `$`, `[]`, `[[]]`, and matrix-style subsetting to access and modify data frame contents.
- Explain the differences between the four subsetting operators.
- Explain how recycling applies when modifying data frame columns.

Take the [Knowledge Quiz](/quizzes/data_frames/index.qmd) to evaluate your learning.

:::
