Question
========

```{r, echo = FALSE, include = FALSE}
l <- list(name = "bob",
          age = 25,
          height = 180,
          hair = "blue")
```

There is a problem with the list shown below (`l`): the information for eye color is wrongly saved in a slot called
`hair`. Which of the following options will change the name of the `hair` slot to something more appropriate?

```{r}
#| echo: false
l
```


Answerlist
----------

* `names(l)[4] <- "eyes"`
* `l$hair <- names("eyes")`
* `l[[4]] <- "eyes"`
* `l[4] <- "eyes"`
* `names(l)$hair <- "eyes"`

Solution
========

The correct way to change the name of the value is `names(l)[4] <- "eyes"`

Answerlist
----------

* Correct
* Wrong: `names("eyes")` is `NULL` since a character string doesn't have a name, so this command will just erase the
list element
* Wrong: This command will change the value itself and not the name
* Wrong: This command will change the value itself and not the name
* Wrong: This command will return an error since the `names()` element is a vector and not a list

Meta-information
================

exname: Default Arguments Selection
extype: schoice
exsolution: 10000
exshuffle: 5
