---
title: "Creating Lists"
---

Lists place essentially no restrictions on the nature of their contents:

- A list slot can hold any valid R object (including other lists).
- The contents of different list slots do not need to have the same type (nor do they need to be different types).
- The contents of different list slots do not need to have the same---or otherwise conformable---sizes.

So, lists are a great way to store substantively related, but heterogeneously sized/typed, pieces of data in one object.

## Generating Lists

We construct lists using the `list()` function.

```{r}
(l1 <- list(1, 2, 3))
(l2 <- list("bob", TRUE, 33, 42 + 3i))
(l3 <- list(c(1, 2), c(3, 4)))
```

::: practice

Create a list called `myList` that comprises the following three elements.

1. A length-three numeric vector
1. A length-one character vector
1. A length-two logical vector

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

Here's one possible solution.

```{r}
myList <- list(
  1:3,
  "alice is not bob",
  rep(TRUE, 2)
)

myList
```

::::

:::

## Naming List Elements

We can name the elements of a list by defining the list slots via key-value pairs.

```{r}
l4 <- list(
  name = "bob",
  alive = TRUE,
  age = 33,
  relationshipStatus = 42 + 3i
)

l4
```

We can use the `names()` function to access and modify the names of a list's slots.

```{r}
# The slots in l1 don't yet have names
names(l1)
l1

# Add names to the l1 slots
names(l1) <- c("first", "second", "third")
l1
```

The `names()` function returns a character vector. So, we can assign/overwrite a subset of the names using ordinary
vector subsetting syntax.

```{r}
# Change the name for the second slot in l4
names(l4)[2] <- "not dead"
names(l4)
l4
```

::: practice

Consider the following list (`aList`) when answering this question.

```{r}
#| echo: false

aList <- list(
	hello = "world",
	day = "monday",
	name = "jude",
	false = TRUE
)
```

```{r}
aList
```

:::: {.panel-tabset}

## {{< var tab_title.question >}}

**Predict** what the following code will print.

```{r}
#| eval: false

names(aList)[2:3] <- c("blue", "hey")
aList
```

Use the interactive editor to check your predictions.

## {{< var tab_title.editor >}}

```{webr-r}
#| context: setup

aList <- list(
	hello = "world",
	day = "monday",
	name = "jude",
	false = TRUE
)
```

```{webr-r}
```

## {{< var tab_title.solution >}}

We've overwritten the names of the second and third list slots with the new values `blue` and `hey`, respectively. So,
we see an updated version of `aList` that contains the same data but has two new names.

```{r}
#| echo: false

names(aList)[2:3] <- c("blue", "hey")
aList
```

::::

:::

## Nested Lists

It's difficult to overstate the versatility of lists as general-purpose containers for R objects. It's no exaggeration
to say that a list can store any type of R object. Notably, lists are perfectly happy to store other lists. So the
following abomination is a completely valid list.

```{r}
level0 <- list(
	level1 = list(
		level2a = list(
			level3a = list("How much wood"),
			level3b = list("would a woodchuck chuck"),
			level3c = list("if")
		),
		level2b = list(
			level3a = list("a"),
			level3b = list("woodchuck could"),
			level3c = list("chuck wood?")
		)
	)
)
```

::: practice

Create a list to describe yourself. Include the following named elements in your list:

1. Your Name
1. Your Eye Color
1. Your Hair Color
1. Your Favorite Color

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

```{r}
myInfo <- list(
  name = "bob",
  eyes = "blue",
  hair = "black",
  color = "green"
)

myInfo
```

::::

:::

::: knowledge-check

- What's the key difference between a list slot and a vector element, in terms of what kind of data each can hold?
- If you don't name a list's slots when you generate the list, can you still add names later later? How?

:::
