---
title: "Manipulating Lists"
---

Like vectors and matrices, we can access and modify the slots of a list. Unlike those simpler data objects, however, we
have several different ways to select list elements (not just the `[]` operator).

## Accessing List Slots

R provides several mechanisms for accessing the contents of a list, each serving a distinct purpose. The three main
operators used for this task are the dollar sign (`$`), single brackets (`[]`), and double brackets (`[[]]`).

### Accessing Single Slots

The `$` operator accesses *a single list slot* by name and returns *the contents* of that list slot. 

```{r}
l1 <- list(
  name = "bob",
  alive = TRUE,
  age = 33,
  relationshipStatus = 42 + 3i
)

l1$name
class(l1$name)
```

The `[[]]` operator works like the `$` operator but also accepts numeric indices.

```{r}
l1[["age"]]
l1[[2]]
class(l1[[2]])
```

### Multi-Slot Access

The `[]` operator is notably different: it extracts *any number of list slots* and returns the result *as a list*. Even
if we only select one element, the `[]` operator still returns a list. This behavior is useful when we want to retain
the list structure for further processing.

```{r}
l1["name"]
l1[2]
class(l1[2])

l1[c("name", "alive")]
l1[1:2]
class(l1[1:2])
```

If we want to select multiple list elements, the `[]` operator is our only option. You cannot use the `$` or `[[]]`
operators to select multiple list slots.

```{r}
#| error: true

l1[[1:2]]
l1$c("name", "alive")
```

::: {.callout-tip}

## A Mental Model for `$`, `[[]]`, and `[]`

To gain some intuition for the different flavors of list subsetting, you can think of a list as a stack of moving boxes.
Let's consider a three-slot list as our example. We conceptualize this list as a stack of three boxes. Let's say the
first box contains clothes, the second box contains books, and the third box contains four smaller boxes that hold
various types of office supplies.

- The `$` and `[[]]` operators work like reaching *inside* a single box and retrieving its *contents*.
  - In our analogy, these two operators unpack a single box.
	- We remove the clothes from our first box and do something with those clothes.
- The `[]` operator works like choosing one or more boxes from our stack and doing something with those unopened boxes.
  - In our analogy, this operator picks up some boxes but doesn't open them.
	- We pick up the first box and move it to our bedroom.

If you ever find yourself surprised that a selection "is still a list" when you expected the raw value from the list
slot, check whether you used `[]` (which always keeps the list wrapper) instead of `$` or `[[]]` (which unwrap it).

:::

**Predict:** What type of object will the following code print?

```{webr-r}
someList <- list(
  a = 1:10,
  b = tail(LETTERS, 5),
  c = list(
    x = "the quick brown",
    y = "fox",
    z = "jumped over"
  )
)

someList$c
```

::: explanation

```{r}
#| include: false

someList <- list(
  a = 1:10,
  b = tail(LETTERS, 5),
  c = list(
    x = "the quick brown",
    y = "fox",
    z = "jumped over"
  )
)
```

```{r}
class(someList$c)
```

The command `someList$c` returns a list because the `c` slot in `someList` holds an entire list. Using the single
element extraction procedure returns the contents of the `c` slot, which is another list, in this case.

So, don't fall into the trap of thinking that `$` and `[[]]` *never* return a list. Both the `$` and `[[]]` operators
will return a list, when the selected slot holds a list.

:::

## Modifying List Elements

As with vectors and matrices, you shouldn't think about the list selection operators (`$`, `[]`, `[[]]`) as "subsetting
operators". It's better to think about the selection operators as designating some part of a list for further
processing. In the above examples, the further processing was simply printing the selected elements to the R console,
but we can also use the selection operators to modify the designated list elements.

```{r}
# View the original list
l1

# Modify some list elements
l1$age <- 57
l1[[1]] <- "suzy"
l1[c(2, 4)] <- c("foo", "bar")

# View the modified list
l1
```

Lists in R do not require a fixed length or predefined structure. New elements can be added at any point, either by
position or by name. This flexibility makes lists well suited for incremental construction.

```{r}
# Create an empty list
(l2 <- list())

# Add new list slots
l2$grass <- "green"
l2$logical <- FALSE
l2[[3]] <- 1:4

l2
```

## Using List Elements

Importantly, the types of the objects you store in a list are maintained.

```{r}
(x <- 1:5)
(y <- list())
(z <- letters[1:10])

(l3 <- list(a = x, b = y, c = z))

class(x)
class(y)
class(z)

class(l3$a)
class(l3$b)
class(l3$c)
```

Consequently, objects stored in a list behave just as they would outside of the list. For example, if a function is
stored as an element of a list, we can call it directly:

```{r}
# Store the Base R 'mean' function in the 'l3' list
l3$myMean <- mean

# Call both versions of our function
mean(1:5)
l3$myMean(1:5)
```

::: practice

Use the list that you created in the last practice problem for this exercise.

:::: {.panel-tabset}

## {{< var tab_title.question >}}

Using a single command, test if your eye color OR your hair color is also your favorite color.

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.hint >}}

Recall the basic logical operators:

- `==`, `!=`: Equal, Not Equal
- `>`, `<`, `>=`, `<=`: Greater Than, Less Than
- `&`, `|`: Logical AND, Logical OR (not exclusive)

## {{< var tab_title.solution >}}

Recall my solution for the last practice problem.

```{r}
#| include: false

myInfo <- list(
  name = "bob",
  eyes = "blue",
  hair = "black",
  color = "green"
)
```

```{r}
myInfo
```

So, I would set up the necessary logical test as follows.

```{r}
myInfo$eyes == myInfo$color | myInfo$hair == myInfo$color
```

::::

:::

::: knowledge-check

- If `l$x` returns an error but `l[["x"]]` works, what does that tell you about `l`?
- Assuming the `name` slot in the list `l` does not hold another list, why would `l["name"]` return a list, while
`l[["name"]]` returns some other type of object?
- Which operator would you need if you wanted to select *two* named list slots at once?

:::

::: {.callout-tip}

## Ready to Continue?

You're ready to move on to [Data Frames](/tutorials/data_frames/index.qmd) if you can:

- Explain what a list is and describe at least one situation where a list is the right tool.
- Create a list with named elements.
- Explain the difference between `$`, `[]`, and `[[]]`, and choose the right one for a given task.

If any of these still feel shaky, revisit the relevant page above, or try the
[Knowledge Quiz](/quizzes/lists/index.qmd) to pinpoint exactly where to focus.

:::
