---
title: "Manipulating Lists"
---

::: {.callout-note}
Using `$`, `[]`, and `[[]]` correctly allows you to apply using lists; being able to explain *why* they behave differently (and predict which one you need in a given situation) is allows you to understand how lists work. This distinction is one of the most common sources of confusion for R beginners, so it's worth extra attention.
:::

## Accessing List Slots

R provides several mechanisms for accessing the contents of a list, each serving a distinct purpose. The three main
operators used for this task are the dollar sign (`$`), single brackets (`[]`), and double brackets (`[[]]`).

The `$` operator accesses *a single list slot* by name and returns *the contents* of that list slot. 

```{r}
l1 <- list(name = "bob",
           alive = TRUE,
           age = 33,
           relationshipStatus = 42+3i)

l1$name
class(l1$name)
```

The `[[]]` operator works like the `$` operator but also accepts numeric indices.

```{r}
l1[["age"]]
l1[[2]]
class(l1[[2]])
```

The `[]` operator is notably different: it extracts *any number of list slots* and returns the result *as a list*.
Even if only one element is selected, the output remains a list. This behavior is useful when we want to retain the list
structure for further processing.

```{r}
l1["name"]
l1[2]
class(l1[2])

l1[c("name", "alive")]
l1[1:2]
class(l1[1:2])
```

If we want to select multiple list elements, the `[]` operator is our only option. You cannot use the `$` or `[[]]`
operators to select multiple list slots.

```{r, error = TRUE}
l1[[1:2]]
l1$c("name", "alive")
```

::: {.callout-tip}

## A Mental Model for `$`, `[[]]`, and `[]`

A common way people remember the difference: think of a list as a row of labeled boxes.

- `$` and `[[]]` reach *inside* a single box and hand you what's in it, the *contents*.
- `[]` hands you a smaller row of boxes, still sealed, the *list itself*, just with fewer slots.

If you ever find yourself surprised that a selection "is still a list" when you expected the raw value, check
whether you used `[]` (which always keeps the list wrapper) instead of `$` or `[[]]` (which unwrap it).

:::

## Modifying List Elements

Above, we used the three selection operators to extract list elements, but you shouldn't think about those operators as
"subsetting operators". It's better to think about the selection operators as designating some part of a list for further
processing. In the above examples, the further processing was simply printing the selected elements to the R console, but
we can also use the selection operators to modify the designated list elements.

```{r}
# View the original list
l1

## Modify some list elements
l1$age <- 57
l1[[1]] <- "suzy"
l1[c(2, 4)] <- c("foo", "bar")

# View the modified list
l1
```

Lists in R do not require a fixed length or predefined structure. New elements can be added at any point, either by
position or by name. This flexibility makes lists well suited for incremental construction.

```{r}
# Create an empty list
(l2 <- list())

# Add new list slots
l2$grass <- "green"
l2$logical <- FALSE
l2[[3]] <- 1:4

l2
```

Importantly, the types of the objects you store in a list are maintained.

```{r}
(x <- 1:5)
(y <- list())
(z <- letters[1:10])

(l3 <- list(a = x, b = y, c = z))

class(x)
class(y)
class(z)

class(l3$a)
class(l3$b)
class(l3$c)
```

Consequently, objects stored in a list behave just as they would outside of the list. For example, if a function is
stored as an element of a list, we can call it directly:

```{r}
# Create a trivial function
addOne <- function(x) { x + 1 }

# Store our function object in the 'l3' list
l3$myFunction <- addOne

# Call both versions of our function
addOne(2)
l3$myFunction(2)
```

::: practice

## Practice

Use the list that you created in the last practice problem for this exercise.

Using a single command, test if your eye color OR your hair color is also your favorite color.

:::: {.panel-tabset}

## {{< var tab_title.question >}}

```{webr-r}

```

## {{< var tab_title.hint >}}

Recall the basic logical operators:

- `==`, `!=`: Equal, Not Equal
- `>`, `<`, `>=`, `<=`: Greater Than, Less Than
- `&`, `|`: Logical AND, Logical OR (not exclusive)

## {{< var tab_title.solution >}}

```{r, echo = FALSE, include = FALSE}
myInfo <- list(name = "bob",
               eyes = "blue",
               hair = "black",
               color = "green")
```

```{r}
myInfo$eyes == myInfo$color | myInfo$hair == myInfo$color
```

::::

:::

## Quick Check

- If `l$x` returns an error but `l[["x"]]` works, what does that tell you about `l`?
- Why does `l["name"]` return a list, while `l[["name"]]` returns just the value?
- Which operator would you need if you wanted to select *two* named slots at once? Why can't you use `$` or `[[]]`
  for that?

::: {.callout-tip}

## Ready to Continue?

You're ready to move on to [Data Frames](/tutorials/data_frames/index.qmd) if you can:

- Explain what a list is and describe at least one situation where a list is the right tool.
- Create a list with named elements.
- Explain the difference between `$`, `[]`, and `[[]]`, and choose the right one for a given task.

If any of these still feel shaky, revisit the relevant page above, or try the
[Knowledge Quiz](/quizzes/lists/index.qmd) to pinpoint exactly where to focus.

:::
